---
title: "01 Linear-Truth Cross-Validation Replication"
author: Aileen Tao
format:
  pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = NA_character_)
if (is.na(root) || !nzchar(root)) {
  here::i_am("report/01_cv_linear_truth.qmd")
  root <- here::here()
}
options(`_01_cv_linear_truth_report_only` = TRUE)
source(file.path(root, "code", "01_run_cv_linear_truth.R"), local = knitr::knit_global())
options(`_01_cv_linear_truth_report_only` = NULL)
```

# Goal

This project simulates data under a linear truth and compares two candidate models with one random 10-fold split per Monte Carlo replication:

- Linear model (correct): `y ~ x`
- Polynomial model: `y ~ poly(x, 3)`

The aim is to show that CV model selection still varies across finite replications.

# Simulation settings

```{r parameter-check, echo=FALSE}
knitr::kable(
  settings_tbl,
  caption = "Simulation settings."
)
```

# Selection Counts and Rates

```{r selection-summary, echo=FALSE}
knitr::kable(
  selection_summary_tbl,
  digits = 4,
  caption = "How often each model is selected across 1000 replications for each sample size."
)
```

# CV Error Summary

```{r gap-summary, echo=FALSE}
knitr::kable(
  gap_summary_tbl,
  digits = 5,
  caption = "Average CV error and gap summary by sample size (poly3 minus linear)."
)
```

# Conclusion

Even when the true model is linear, cross-validation can select either linear or polynomial in finite replications. The selection-count table quantifies this variation for both large-sample datasets (`n = 10000` and `n = 30000`).
