---
title: "01 Linear-Truth Cross-Validation Replication"
author: Aileen Tao
format:
  pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
```

# Goal

This project simulates data under a linear truth and compares two candidate models with one random 10-fold split per Monte Carlo replication:

- Linear model (correct): `y ~ x`
- Polynomial model: `y ~ poly(x, 3)`

The aim is to show that CV model selection still varies across finite replications.

# Load Results

```{r load-results}
rds_candidates <- c(
  "results/01_cv_linear_truth_results.rds",
  "report/results/01_cv_linear_truth_results.rds",
  "../report/results/01_cv_linear_truth_results.rds"
)
rds_found <- rds_candidates[file.exists(rds_candidates)]

if (length(rds_found) > 0) {
  output <- readRDS(rds_found[[1]])
  params <- output$parameters
  results <- output$results
  selection_summary <- output$selection_summary
  gap_summary <- output$gap_summary
} else {
  stop(
    "Missing results file: report/results/01_cv_linear_truth_results.rds. ",
    "Run code/01_run_cv_linear_truth.R first."
  )
}
```

```{r parameter-check, echo=FALSE}
knitr::kable(
  data.frame(
    setting = c("K", "Monte Carlo replications per sample size", "Sample sizes", "beta0", "beta1", "sigma", "poly_degree"),
    value = c(params$K, params$mc_runs, paste(params$sample_sizes, collapse = ", "), params$beta0, params$beta1, params$sigma, params$poly_degree)
  ),
  caption = "Simulation settings."
)
```

# Selection Counts and Rates

```{r selection-summary, echo=FALSE}
knitr::kable(
  selection_summary |>
    arrange(n, selected_model),
  digits = 4,
  caption = "How often each model is selected across 1000 replications for each sample size."
)
```

# CV Error Summary

```{r gap-summary, echo=FALSE}
knitr::kable(
  gap_summary |>
    arrange(n),
  digits = 5,
  caption = "Average CV error and gap summary by sample size (poly3 minus linear)."
)
```

# Conclusion

Even when the true model is linear, cross-validation can select either linear or polynomial in finite replications. The selection-count table quantifies this variation for both large-sample datasets (`n = 10000` and `n = 30000`).
