---
title: "02 CV Selection Error Rate vs Sample Size"
author: Aileen Tao
format:
  pdf:
    fig-width: 7
    fig-height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = NA_character_)
if (is.na(root) || !nzchar(root)) {
  here::i_am("report/02_cv_error_rate.qmd")
  root <- here::here()
}
options(cv_error_rate_02_report_only = TRUE)
source(file.path(root, "code", "02_cv_error_rate.R"), local = knitr::knit_global())
options(cv_error_rate_02_report_only = NULL)
```

# Goal

Data are simulated under a **linear** truth. For each sample size on a grid from 50 to 10,000 (step 50), we run 10-fold cross-validation for a **linear** model and a **degree-3 polynomial** model, pick the model with lower mean CV MSE, and record whether CV incorrectly chose the polynomial. The **CV selection error rate** is the fraction of 1,000 Monte Carlo replications where polynomial wins.

# Simulation settings

```{r settings, echo=FALSE}
knitr::kable(
  settings_tbl,
  caption = "Simulation and CV settings for analysis 02."
)
```

# CV error summary by sample size

```{r error-summary, echo=FALSE}
knitr::kable(
  error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is linear; error = poly chosen) and approximate MC standard error."
)
```

# CV error rate vs sample size

```{r error-plot, echo=FALSE, fig.cap = "CV selection error rate as a function of sample size (ribbon = MC SE band, clipped to [0, 1])."}
print(error_rate_plot)
```

