---
title: "03 Rolling Validation Selection Error Rate (Expanding Window)"
author: Aileen Tao
format:
  pdf:
    fig-width: 7
    fig-height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = NA_character_)
if (is.na(root) || !nzchar(root)) {
  here::i_am("report/03_expanding_rolling_error_rate.qmd")
  root <- here::here()
}
source(file.path(root, "code", "03_rv_expanding_linear.R"), local = knitr::knit_global())
source(file.path(root, "code", "03_rv_expanding_poly.R"), local = knitr::knit_global())
```

# Goal (expanding window)

One-step-ahead **expanding** rolling validation: initial training window `w = floor(0.5 * n)`, then for `t = w, …, n - 1` fit on rows `1:t` and predict row `t + 1`. Mean squared prediction error is averaged over `n - w` steps. We compare a **linear** model (M1) and **degree-3 polynomial** (M3), select the model with lower mean RV MSE (tie → linear), and report the Monte Carlo **selection error rate** vs `n` on `50, 100, …, 1000` (step 50), with 1,000 replications per `n`.

# Linear truth

## Simulation settings

```{r expanding-linear-simulation-settings, echo=FALSE}
knitr::kable(
  expanding_linear_simulation_settings_tbl,
  caption = "Simulation settings (expanding window, linear truth)."
)
```

## Error rate summary

```{r expanding-linear-summary, echo=FALSE}
knitr::kable(
  expanding_linear_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is linear; error = poly chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r expanding-linear-plot, echo=FALSE, fig.cap = "RV selection error rate under linear truth (expanding window)."}
print(expanding_linear_error_rate_plot)
```

# Polynomial truth

## Simulation settings

```{r expanding-poly-simulation-settings, echo=FALSE}
knitr::kable(
  expanding_poly_simulation_settings_tbl,
  caption = "Simulation settings (expanding window, polynomial truth)."
)
```

## Error rate summary

```{r expanding-poly-summary, echo=FALSE}
knitr::kable(
  expanding_poly_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is polynomial; error = linear chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r expanding-poly-plot, echo=FALSE, fig.cap = "RV selection error rate under polynomial truth (expanding window)."}
print(expanding_poly_error_rate_plot)
```
