---
title: "03 Rolling Validation Selection Error Rate (Sliding Window)"
author: Aileen Tao
format:
  pdf:
    fig-width: 7
    fig-height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = NA_character_)
if (is.na(root) || !nzchar(root)) {
  here::i_am("report/03_sliding_rolling_error_rate.qmd")
  root <- here::here()
}
source(file.path(root, "code", "03_sliding_linear.R"), local = knitr::knit_global())
source(file.path(root, "code", "03_sliding_poly.R"), local = knitr::knit_global())
```

# Goal (sliding window)

One-step-ahead **sliding** rolling validation: fixed window `w = floor(0.5 * n)`; for `t = 1, …, n - w` fit on rows `t:(t + w - 1)` and predict row `t + w`. Mean squared prediction error is averaged over `n - w` steps (same step count as expanding at the same `n` and `train_frac`). Model comparison and grid match the expanding-window analysis.

# Linear truth

## Simulation settings

```{r sliding-linear-settings, echo=FALSE}
knitr::kable(
  sliding_linear_settings_tbl,
  caption = "Rolling validation settings (sliding window, linear truth)."
)
```

## Error rate summary

```{r sliding-linear-summary, echo=FALSE}
knitr::kable(
  sliding_linear_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is linear; error = poly chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r sliding-linear-plot, echo=FALSE, fig.cap = "RV selection error rate under linear truth (sliding window)."}
print(sliding_linear_error_rate_plot)
```

# Polynomial truth

## Simulation settings

```{r sliding-poly-settings, echo=FALSE}
knitr::kable(
  sliding_poly_settings_tbl,
  caption = "Rolling validation settings (sliding window, polynomial truth)."
)
```

## Error rate summary

```{r sliding-poly-summary, echo=FALSE}
knitr::kable(
  sliding_poly_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is polynomial; error = linear chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r sliding-poly-plot, echo=FALSE, fig.cap = "RV selection error rate under polynomial truth (sliding window)."}
print(sliding_poly_error_rate_plot)
```
