---
title: "04 Cross-Validation Selection Error Rate"
author: Aileen Tao
execute:
  dir: project
format:
  pdf:
    fig-width: 7
    fig-height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
here::i_am("report/04_cv_error_rate.qmd")
root <- here::here()
source(file.path(root, "code", "04_cv_error_rate_linear.R"), local = knitr::knit_global())
source(file.path(root, "code", "04_cv_error_rate_poly.R"), local = knitr::knit_global())
```

# Goal (cross-validation)

For each sample size on `50, 100, …, 1000` (step 50), we run **10-fold** cross-validation for a **linear** model (M1) and **degree-3 polynomial** (M3), select the model with lower mean CV MSE (tie → linear), and report the Monte Carlo **selection error rate** vs `n`, with **500** replications per `n`.

# Linear truth

## Simulation settings

```{r cv-linear-simulation-settings, echo=FALSE}
knitr::kable(
  cv_linear_simulation_settings_tbl,
  caption = "Simulation settings (cross-validation, linear truth)."
)
```

## Error rate summary

```{r cv-linear-summary, echo=FALSE}
knitr::kable(
  cv_linear_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is linear; error = poly chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r cv-linear-plot, echo=FALSE, fig.cap = "CV selection error rate under linear truth."}
print(cv_linear_error_rate_plot)
```

# Polynomial truth

## Simulation settings

```{r cv-poly-simulation-settings, echo=FALSE}
knitr::kable(
  cv_poly_simulation_settings_tbl,
  caption = "Simulation settings (cross-validation, polynomial truth)."
)
```

## Error rate summary

```{r cv-poly-summary, echo=FALSE}
knitr::kable(
  cv_poly_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is polynomial; error = linear chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r cv-poly-plot, echo=FALSE, fig.cap = "CV selection error rate under polynomial truth."}
print(cv_poly_error_rate_plot)
```
