---
title: "04 Rolling Validation Selection Error Rate (Fixed Window)"
author: Aileen Tao
format:
  pdf:
    fig-width: 7
    fig-height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = NA_character_)
if (is.na(root) || !nzchar(root)) {
  here::i_am("report/04_fixed_rolling_error_rate.qmd")
  root <- here::here()
}
source(file.path(root, "code", "04_rv_fixed_linear.R"), local = knitr::knit_global())
source(file.path(root, "code", "04_rv_fixed_poly.R"), local = knitr::knit_global())
```

# Goal (fixed window)

One-step-ahead **fixed** rolling validation with training window **`w = 50`**: for `t = 1, …, n - w`, fit on rows `t:(t + w - 1)` and predict row `t + w`. Mean squared prediction error is averaged over `n - w` steps. We compare a **linear** model (M1) and **degree-3 polynomial** (M3), select the model with lower mean RV MSE (tie → linear), and report the Monte Carlo **selection error rate** vs `n` on **`100, 150, …, 1000`** (step 50), with 1,000 replications per `n`.

# Linear truth

## Simulation settings

```{r fixed-linear-simulation-settings, echo=FALSE}
knitr::kable(
  fixed_linear_simulation_settings_tbl,
  caption = "Simulation settings (fixed window w = 50, linear truth)."
)
```

## Error rate summary

```{r fixed-linear-summary, echo=FALSE}
knitr::kable(
  fixed_linear_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is linear; error = poly chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r fixed-linear-plot, echo=FALSE, fig.cap = "RV selection error rate under linear truth (fixed window w = 50)."}
print(fixed_linear_error_rate_plot)
```

# Polynomial truth

## Simulation settings

```{r fixed-poly-simulation-settings, echo=FALSE}
knitr::kable(
  fixed_poly_simulation_settings_tbl,
  caption = "Simulation settings (fixed window w = 50, polynomial truth)."
)
```

## Error rate summary

```{r fixed-poly-summary, echo=FALSE}
knitr::kable(
  fixed_poly_error_rate_summary_tbl,
  digits = 5,
  caption = "Monte Carlo mean selection-error rate (truth is polynomial; error = linear chosen) and approximate MC standard error."
)
```

## Error rate vs sample size

```{r fixed-poly-plot, echo=FALSE, fig.cap = "RV selection error rate under polynomial truth (fixed window w = 50)."}
print(fixed_poly_error_rate_plot)
```
