---
title: "07 Slowly Varying Weights for Rolling Validation"
author: Aileen Tao
execute:
  dir: project
format:
  pdf:
    fig-width: 7
    fig-height: 4.5
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{placeins}
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
              fig.pos = "H", out.width = "100%")
here::i_am("report/07_slowly_varying_rv.qmd")
root <- here::here()
source(file.path(root, "code", "07_slowly_varying_rv.R"), local = knit_global())
```

# Research question and choice of weights

Analysis 06 studied power weights $w_i=i^\alpha$. Here I ask whether
$\log i$, a polynomial of $\log i$, $\log\log i$, and slowly decreasing
weights lead to consistent selection in the same Gaussian example.

## Candidate weight functions

The seven main weights cover a baseline, slowly increasing weights, and
two decreasing weights.

```{r candidate-table}
candidates07 <- data.frame(
  Weight = c("$1$", "$\\log(i+1)$", "$[\\log(i+1)]^2$",
             "$1+\\log(i+1)+[\\log(i+1)]^2$", "$\\log\\log(i+e)$",
             "$1/\\log(i+1)$", "$1/[\\log(i+1)]^2$"),
  Reason = c("Unweighted baseline", "Check the log idea",
             "Higher log power", "Check a polynomial of log",
             "Check the log-log idea", "Test decreasing weights",
             "Test a faster decrease"),
  Role = c("Reference", "Slow growth", "Stronger slow growth",
           "Genuine polynomial of log", "Extremely slow growth",
           "Possible boundary case", "Compare the ratio criterion"),
  check.names = FALSE)
kable(candidates07, format = "latex", booktabs = TRUE, escape = FALSE,
      col.names = c("Weight", "Why test it?", "Expected role")) |>
  kable_styling(latex_options = "HOLD_position", font_size = 8) |>
  column_spec(1, width = "5.2cm") |>
  column_spec(2, width = "4.0cm") |>
  column_spec(3, width = "3.8cm")
```

## Why use shifted logarithms?

The shifts keep every weight defined and positive at small indices:
$1/\log1$ is undefined, while $i+e>e$ implies
$\log\log(i+e)>0$. They do not change the asymptotics because
$\log(i+1)\sim\log i$ and $\log\log(i+e)\sim\log\log i$.

The family $w_i=[\log(i+1)]^p$ contains five of the main cases.
I also examine $p=-1.25,-0.75,-0.5,0.5$ near the boundary and include
$i^{0.01}$ as a reference to Analysis 06.

# Common setup and decision criterion

## What counts as a wrong selection?

Let $Y_i=\theta_\star+\xi_i$, where the $\xi_i$ are independent $N(0,1)$,
and put $S_m=\sum_{t=1}^m\xi_t$. At time $i$, compare
$\widehat\theta_{i-1}=(i-1)^{-1}\sum_{t<i}Y_t$ with the oracle
$\theta_\star$. For positive deterministic weights $w_i$, their weighted
squared-loss difference is

$$
\Xi_n=\sum_{i=2}^n w_i
\left\{\frac{S_{i-1}^2}{(i-1)^2}
-2\xi_i\frac{S_{i-1}}{i-1}\right\}.
\tag{1}
$$

A positive score favors the oracle; $\Xi_n<0$ is a wrong selection.

## What the Chebyshev ratio can establish

Write

$$
M_n=\mathbb E\Xi_n,\qquad V_n=\operatorname{Var}(\Xi_n),
\qquad R_n=\frac{V_n}{M_n^2}.
$$

Since $M_n>0$, Chebyshev's inequality gives

$$
\mathbb P(\Xi_n<0)\leq \min\{1,R_n\}.
\tag{2}
$$

Thus $R_n\to0$ proves consistency. A positive ratio limit is inconclusive;
it does not prove inconsistency.

## One formula for all candidate weights

The following formula applies to every deterministic weight considered here.

**Proposition 1.** For any positive deterministic weights, define

$$
\begin{aligned}
M_n&=\sum_{i=2}^n\frac{w_i}{i-1},\\
a_{n,t}&=\sum_{i=t+1}^n\frac{w_i}{(i-1)^2},
&&1\leq t<n,\\
b_{n,l}&=\sum_{i=l+1}^n\frac{w_i}{(i-1)^2}
          -\frac{w_l}{l-1},
&&2\leq l\leq n.
\end{aligned}
$$

Then

$$
\begin{aligned}
\Xi_n-M_n
 &=\sum_{t=1}^{n-1}a_{n,t}(\xi_t^2-1)
   +2\sum_{l=2}^n b_{n,l}\sum_{k<l}\xi_k\xi_l,\\
V_n
 &=2\sum_{t=1}^{n-1}a_{n,t}^2
   +4\sum_{l=2}^n(l-1)b_{n,l}^2.
\end{aligned}
\tag{3}
$$

*Proof.* The sample mean at time $i$ uses only earlier observations, so
$S_{i-1}$ is independent of $\xi_i$. This gives
$\mathbb E(\xi_i S_{i-1})=0$ and
$\mathbb E S_{i-1}^2=i-1$, yielding the mean.
To obtain the variance, expand $S_{i-1}^2$ in (1) and gather the terms
containing each $\xi_t^2$ and each $\xi_k\xi_l$.
The centered squares have variance $2$, and each product $\xi_k\xi_l$
has variance $1$. Distinct centered squares are uncorrelated.
A square and a product are uncorrelated because some independent centered
Gaussian variable appears to an odd power. The same reasoning applies to two
different products, including products sharing one index. There are $l-1$
products with larger index $l$, so summing their variances gives (3).
$\square$

For the weights considered here, it helps to separate coefficients that
would remain if validation continued indefinitely from the change caused
by stopping at $n$. Suppose
$\sum_{i=2}^\infty w_i/(i-1)^2<\infty$, and define

$$
A_t=\sum_{i=t+1}^\infty\frac{w_i}{(i-1)^2},
\qquad B_l=A_l-\frac{w_l}{l-1},
\qquad T_n=A_n.
$$

Here $A_t$ is the infinite future tail of a diagonal coefficient,
$T_n=A_n$ is the tail removed by stopping at $n$, and $B_l$ is the
off-diagonal residual after subtracting the direct validation contribution.
The quantity $B_l$ is important because its leading terms cancel.

Then $a_{n,t}=A_t-T_n$ and $b_{n,l}=B_l-T_n$ exactly. Since the weights are
positive, $0\leq a_{n,t}\leq A_t$. Using $(x-y)^2\leq2x^2+2y^2$ in (3) gives

$$
V_n\leq
2\sum_{t<n}A_t^2+
8\sum_{l=2}^n(l-1)B_l^2+
4n(n-1)T_n^2.
\tag{4}
$$

The first sum is the diagonal contribution; the other terms bound the
off-diagonal contribution. **For the log-power weights, the diagonal part
stays bounded.**

If the sum of
$\sum_{t<n}A_t^2$, $\sum_{l=2}^n(l-1)B_l^2$, and $n^2T_n^2$
is $o(M_n^2)$, then $R_n\to0$ and selection is consistent.

**Key point.** This condition does not require increasing weights; it only
requires the mean signal to dominate the variance. This is why decreasing
weights are included.

# What happens in the log-power family?

Consider $w_i=[\log(i+1)]^p$, with fixed $p\in\mathbb R$, and write
$L=\log n$.

**Proposition 2.** For the log-power family, the ratio tends to zero exactly
when $p\geq-1$.
For decreasing weights, I use two limiting constants:
$M_\infty(p)=\sum_{i=2}^\infty[\log(i+1)]^p/(i-1)$
is used only for $p<-1$, and, for $p<0$,

$$
V_\infty(p)=2\sum_{t=1}^\infty A_t^2
             +4\sum_{l=2}^\infty(l-1)B_l^2\in(0,\infty).
$$

The first is a finite mean for $p<-1$. The second is the finite variance
reached when $p<0$.

## Roadmap

The mean and variance each have three regimes:

$$
M_n\begin{cases}
\sim L^{p+1}/(p+1),&p>-1,\\[2pt]
\sim\log L,&p=-1,\\[2pt]
\to M_\infty(p),&p<-1,
\end{cases}
\qquad
V_n\begin{cases}
\sim2L^{2p},&p>0,\\[2pt]
\to\pi^2,&p=0,\\[2pt]
\to V_\infty(p),&p<0.
\end{cases}
$$

**The mean has three regimes. The variance has three regimes.**
Because their breakpoints differ, the ratio has five cases:
$p>0$, $p=0$, $-1<p<0$, $p=-1$, and $p<-1$.

## Case 1: $p>0$ -- increasing log powers

The tested log and squared-log weights correspond to $p=1$ and $p=2$.
In this case,

$$
M_n\sim\frac{L^{p+1}}{p+1},\qquad V_n\sim2L^{2p},
\qquad R_n\sim\frac{2(p+1)^2}{L^2}\to0.
$$

The ratio vanishes, so selection is consistent.

## Case 2: $p=0$ -- the unweighted reference

Here $w_i=1$, $M_n\sim L$, and
$V_n=6\sum_{m=1}^{n-1}m^{-2}\to\pi^2$. Consequently,
$R_n\sim\pi^2/L^2\to0$. The mean diverges while the variance stays
bounded, so the baseline is consistent.

## Case 3: $-1<p<0$ -- slowly decreasing weights

For example, $p=-1/2$ gives $w_i=1/\sqrt{\log(i+1)}$. These weights
decrease to zero, yet

$$
M_n\sim\frac{L^{p+1}}{p+1}\to\infty,\qquad
V_n\to V_\infty(p),\qquad
R_n\sim\frac{V_\infty(p)(p+1)^2}{L^{2p+2}}\to0.
$$

The mean diverges and the variance stays bounded. **Decreasing weights
can still be consistent.**

## Case 4: $p=-1$ -- the boundary case

For $w_i=1/\log(i+1)$,

$$
M_n\sim\log L=\log\log n,\qquad V_n\to V_\infty(-1),
\qquad R_n\sim\frac{V_\infty(-1)}{(\log L)^2}\to0.
$$

The mean diverges only as $\log\log n$, so the ratio vanishes slowly.
This boundary case is still consistent.

## Case 5: $p<-1$ -- a positive ratio limit

The main comparison $p=-2$ gives $w_i=1/[\log(i+1)]^2$. Now
$M_n\to M_\infty(p)>0$ and $V_n\to V_\infty(p)>0$, so
$R_n\to V_\infty(p)/M_\infty(p)^2>0$.
The total mean signal stops growing and the variance also has a finite
positive limit. **A positive ratio limit does not prove inconsistency.**

## The boundary in one statement

$$
\boxed{p=-1\text{ is the Chebyshev-ratio boundary for }[\log(i+1)]^p.}
\tag{5}
$$

Specifically, $p\geq-1\Rightarrow R_n\to0$, while
$p<-1\Rightarrow R_n\to\text{a positive constant}$.
On the first side, (2) proves consistency. On the second, the ratio
provides no conclusion about the limiting wrong-selection probability.

**Why $p=-1$?** For $p\geq-1$ the mean still diverges; for $p<-1$ it
converges. Meanwhile, the variance is bounded for $p<0$, and for $p>0$
it grows more slowly than $M_n^2$. This places the ratio boundary at
$p=-1$. As in Analysis 06, a positive ratio limit only makes the
Chebyshev argument inconclusive.

## Why the five cases have these limits

*Proof.* The mean is the easiest part: each validation time contributes
$w_i/(i-1)$. Replacing
$[\log(i+1)]^p/(i-1)$ by $(\log i)^p/i$ changes the partial sum by a bounded
amount: the termwise error is
$O(((\log i)^p+(\log i)^{p-1})/i^2)$, which is summable.
The integral $\int(\log x)^p\,dx/x$, with $u=\log x$, then shows exactly
when those contributions add up without bound:

$$
\begin{aligned}
M_n&\sim L^{p+1}/(p+1), &&p>-1,\\
M_n&\sim\log L, &&p=-1,\\
M_n&\longrightarrow M_\infty(p)\in(0,\infty), &&p<-1.
\end{aligned}
\tag{6}
$$

For the variance, the leading terms cancel:

$$
A_l\sim\frac{(\log l)^p}{l},
\qquad
\frac{w_l}{l-1}\sim\frac{(\log l)^p}{l},
\qquad
B_l=A_l-\frac{w_l}{l-1}.
$$

Thus $B_l$ is smaller than either term separately. More precisely,

$$
\begin{aligned}
A_t&\sim \frac{(\log t)^p}{t},
&T_n&\sim\frac{L^p}{n},\\
|B_l|&\leq C_p\left\{
  \frac{(\log l)^{p-1}}{l}+
  \frac{(\log l)^p}{l^2}\right\}.
\end{aligned}
\tag{7}
$$

Here and below constants may depend on fixed $p$. To estimate $A_t$,
compare its tail sum with $I_p(t)=\int_t^\infty(\log x)^p x^{-2}\,dx$.
On each unit interval the derivative bound gives a total comparison error
$O((\log t)^p/t^2)$. Integration by parts gives
$I_p(t)=(\log t)^p/t+pI_{p-1}(t)$.
Also $I_q(t)=O((\log t)^q/t)$ for every fixed $q$: substitute $x=te^v$,
so the remaining integral is
$t^{-1}\int_0^\infty(\log t+v)^q e^{-v}\,dv$.
The leading term is $(\log t)^p/t$, proving the first line of (7).

To see the cancellation in $B_l$, replace $(i-1)^{-2}$ in $A_l$ by
$[i(i-1)]^{-1}$; the error is $O((\log l)^p/l^2)$.
Summation by parts gives

$$
\sum_{i=l+1}^\infty\frac{w_i}{i(i-1)}
=\frac{w_{l+1}}{l}
+\sum_{i=l+2}^\infty\frac{w_i-w_{i-1}}{i-1}.
$$

The difference between $w_{l+1}/l$ and $w_l/(l-1)$ is
$O((\log l)^p/l^2+(\log l)^{p-1}/l^2)$.
The mean-value theorem bounds
$|w_i-w_{i-1}|\leq C_p(\log i)^{p-1}/i$.
Applying the preceding integral bound to the remaining sum proves (7).
**The leading term in $B_l$ cancels.**
For $p=0$ this derivative term is zero and $B_l=O(l^{-2})$.

Now suppose $p>0$. The diagonal part is bounded because
$a_{n,t}\leq A_t$ and $\sum_t A_t^2<\infty$ by (7). For the
off-diagonal part, define

$$
D_n=\sum_{l=2}^n(l-1)B_l^2.
$$

The quantity $D_n$ collects the $B_l^2$ contribution. It lets us compare
that contribution with the terminal term involving $T_n^2$.

From (7), the first term in the bound for $B_l$ determines the larger
order:

$$
(l-1)B_l^2
=O\left(\frac{(\log l)^{2p-2}}{l}\right),
\qquad
D_n=O\left(\sum_{l=2}^n
\frac{(\log l)^{2p-2}}{l}\right).
$$

Integral comparison reduces the sum to

$$
\int^n\frac{(\log x)^{2p-2}}{x}\,dx
=\int^{\log n}u^{2p-2}\,du,
\qquad u=\log x.
$$

The integral is bounded, logarithmic, or polynomial according as
$2p-2$ is less than, equal to, or greater than $-1$. Solving
$2p-2=-1$ gives the threshold $p=1/2$. Hence

$$
D_n=
\begin{cases}
O(1),&0<p<1/2,\\[2pt]
O(\log\log n),&p=1/2,\\[2pt]
O(L^{2p-1}),&p>1/2.
\end{cases}
$$

The three-way split is only needed for the common conclusion

$$
\boxed{D_n=o(L^{2p})}.
$$

Next, $T_n\sim L^p/n$ and
$\sum_{l=2}^n(l-1)\sim n^2/2$, so

$$
T_n^2\sum_{l=2}^n(l-1)
\sim\frac{L^{2p}}{n^2}\frac{n^2}{2}
=\frac{L^{2p}}{2}.
$$

After the factor $4$ in (3), the terminal contribution is

$$
4T_n^2\sum_{l=2}^n(l-1)\sim2L^{2p}.
$$

Thus the terminal term has order $L^{2p}$, larger than $D_n$.

Since

$$
(B_l-T_n)^2=B_l^2-2B_lT_n+T_n^2,
$$

the cross term satisfies

$$
2\left|T_n\sum_{l=2}^n(l-1)B_l\right|
\leq
2\left\{
\left(\sum_{l=2}^n(l-1)B_l^2\right)
\left(T_n^2\sum_{l=2}^n(l-1)\right)
\right\}^{1/2}.
$$

Equivalently, the right side is
$2\{D_nT_n^2\sum_{l=2}^n(l-1)\}^{1/2}$.
Since $D_n=o(L^{2p})$ and
$T_n^2\sum(l-1)\asymp L^{2p}$, the cross term is
$o(L^{2p})$. Thus Cauchy--Schwarz makes it asymptotically smaller
than the terminal term.

The four variance contributions now have orders

- diagonal part: $O(1)$;
- $D_n$ part: $o(L^{2p})$;
- cross term: $o(L^{2p})$;
- terminal part, including the factor $4$: $\sim2L^{2p}$.

Therefore,

$$
\boxed{V_n\sim2L^{2p},\qquad p>0.}
$$

For $p<0$, $\sum(l-1)B_l^2$ converges, whereas
$n^2T_n^2\sim L^{2p}\to0$. The same cross-term bound tends to zero.
Dominated convergence for the diagonal sum, and convergence of the
$B_l$ series for the off-diagonal sum, give $V_n\to V_\infty(p)$.
This limit is strictly positive because its first term includes $2A_1^2>0$.

At $p=0$, the unit weights give
$V_n=6\sum_{m=1}^{n-1}m^{-2}\to\pi^2$.
For computation, if $W_{n-1}=\sum_{i=2}^{n-1}w_i$, then

$$
V_n-V_{n-1}
=\frac{2w_n}{(n-1)^2}\{(2n-1)w_n-2W_{n-1}\}.
\tag{8}
$$

This recurrence also follows from the quadratic-form matrix: its old block gains
$w_n\mathbf1\mathbf1^\mathsf T/(n-1)^2$, its new off-diagonal entries are
$-w_n/(n-1)$, and the sum of all entries of the old matrix is $-W_{n-1}$.
For unit weights (8) is $6/(n-1)^2$, starting from $V_1=0$.
Combining these variance limits with (6) proves (5). $\square$

Hence

$$
V_n\begin{cases}
\sim2L^{2p},&p>0,\\[2pt]
\to\pi^2,&p=0,\\[2pt]
\to V_\infty(p),&p<0.
\end{cases}
$$

**The variance has three regimes:** the terminal term dominates for
$p>0$, the unit-weight case has a special identity, and the variance
stays bounded for $p<0$.

The case $p=0$ has its own leading constant. For every fixed $p>0$, the
terminal contribution eventually leads; with unit weights it does not.
The positive-$p$ constant therefore cannot simply be evaluated at $p=0$.

# Polynomial and log-log weights

## Polynomial of log

The tested polynomial is

$$
w_i=1+\log(i+1)+[\log(i+1)]^2.
$$

More generally, for
$w_i=P(\log(i+1))$ with $P(x)=cx^k+$ lower-order terms,
the highest-degree term dominates.

**Proposition 3.** Suppose $w_i=P(\log(i+1))>0$, where $P$ is a
polynomial of degree $k\geq1$ with positive leading coefficient $c$.
Then

$$
M_n\sim\frac{c}{k+1}L^{k+1},\qquad
V_n\sim2c^2L^{2k},\qquad
R_n\sim\frac{2(k+1)^2}{L^2}.
\tag{9}
$$

Thus a polynomial of log reduces asymptotically to the log-power case
$p=k$. For the tested degree-two polynomial, $R_n\sim18/L^2\to0$.

*Proof.* The leading term controls the weight because
$P(x)=cx^k+O(x^{k-1})$, so
$P(\log(i+1))=c[\log(i+1)]^k+O([\log(i+1)]^{k-1})$.
When the terms are integrated to obtain the mean, lower degrees contribute
only $O(L^k)$, compared with the leading $cL^{k+1}/(k+1)$.
They also make a smaller contribution to the tail coefficient $A_t$:
$O((\log t)^{k-1}/t)$, lower order than $c(\log t)^k/t$.
The derivative of the weight is $O((\log x)^{k-1}/x)$, so the proof of (7) gives
$A_t\sim c(\log t)^k/t$ and
$B_l=O((\log l)^{k-1}/l+(\log l)^k/l^2)$.
Thus $T_n=A_n\sim cL^k/n$, and $\sum_{l=2}^n(l-1)\sim n^2/2$ gives

$$
4T_n^2\sum_{l=2}^n(l-1)
\sim4\frac{c^2L^{2k}}{n^2}\frac{n^2}{2}
=2c^2L^{2k}.
$$

The lower-degree terms contribute only lower orders to both moments.
The diagonal term is bounded, while
$\sum_{l=2}^n(l-1)B_l^2=O(L^{2k-1})=o(L^{2k})$.
The cross term is also lower order by the Cauchy--Schwarz estimate in
Proposition 2. Equation (3) therefore gives (9), with the highest degree
determining the leading orders. $\square$

## Log-log weight

The exact weight $w_i=\log\log(i+e)$ grows more slowly than $\log i$.

**Proposition 3 (continued).** For this weight,

$$
M_n\sim L\log L,\qquad
V_n\sim2(\log L)^2,\qquad
R_n\sim\frac{2}{L^2}.
\tag{10}
$$

The common $(\log L)^2$ factor cancels from the ratio.

*Proof.* The mean differs by a bounded amount from
$\int^n\log\log x\,dx/x$. With $u=\log x$ and $du=dx/x$,
$\int\log u\,du=u\log u-u$. Hence
$M_n=L\log L-L+O(1)\sim L\log L-L\sim L\log L$, where $L=\log n$.
The derivative of the shifted weight is
$1/((x+e)\log(x+e))$. The same tail comparison and summation by parts give

$$
A_t\sim\frac{\log\log t}{t},
\qquad
B_l=O\left(\frac{1}{l\log l}+\frac{\log\log l}{l^2}\right).
$$

Thus $T_n=A_n\sim\log\log n/n=(\log L)/n$, so

$$
4T_n^2\sum_{l=2}^n(l-1)
\sim4\frac{(\log L)^2}{n^2}\frac{n^2}{2}
=2(\log L)^2.
$$

The sums $\sum A_t^2$ and $\sum(l-1)B_l^2$ stay bounded. The same
Cauchy--Schwarz argument used above makes the cross term smaller than
the terminal term.
Equation (3) therefore gives $V_n\sim2(\log L)^2$, proving (10).
$\square$

Dividing by the squared mean gives

$$
R_n\sim\frac{2(\log L)^2}{L^2(\log L)^2}
=\frac{2}{L^2}=\frac{2}{(\log n)^2}\to0.
$$

Thus the log-log weight is consistent.

## Theory summary for the seven tested weights

Table 1 puts the seven main weights side by side. Here $L=\log n$,
and each finite positive variance limit depends on the particular weight.
The last column records what the Chebyshev ratio establishes.

```{r regime-table}
regimes07 <- data.frame(
  Weight = c("$1$", "$\\log(i+1)$", "$[\\log(i+1)]^2$",
             "$1+\\log(i+1)+[\\log(i+1)]^2$", "$\\log\\log(i+e)$",
             "$1/\\log(i+1)$", "$1/[\\log(i+1)]^2$"),
  Mean = c("$\\sim L$", "$\\sim L^2/2$", "$\\sim L^3/3$",
           "$\\sim L^3/3$", "$\\sim L\\log L$", "$\\sim\\log L$",
           "$\\to M_\\infty(-2)>0$"),
  Variance = c("$\\to\\pi^2$", "$\\sim2L^2$", "$\\sim2L^4$",
               "$\\sim2L^4$", "$\\sim2(\\log L)^2$", "$\\to V_\\infty(-1)>0$",
               "$\\to V_\\infty(-2)>0$"),
  Ratio = c("$\\sim\\pi^2/L^2$", "$\\sim8/L^2$", "$\\sim18/L^2$",
            "$\\sim18/L^2$", "$\\sim2/L^2$",
            "$\\asymp(\\log L)^{-2}$", "$\\to V_\\infty(-2)/M_\\infty(-2)^2>0$"),
  Conclusion = c(rep("Consistent", 6), "Not decided"),
  check.names = FALSE)
kable(regimes07, format = "latex", booktabs = TRUE, escape = FALSE, label = NA,
      col.names = c("Weight", "Mean behavior", "Variance behavior",
                    "Ratio behavior", "Conclusion"),
      caption = "Theory summary for the seven main weights; L = log n.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

# Numerical verification

## Simulation design and checks

The simulations use centered Gaussian observations and the weights listed
above. Exact moments extend to larger $n$; simulation checks finite-sample
agreement.

There are 5,000 independent replications, seed 2026, processed in batches of
250. The sample sizes are
$100,200,500,1000,2000,5000,10000,20000,50000,100000$.
Each Gaussian path is reused across weights and sample sizes, so the
comparisons are dependent.
Exact trajectories use `r length(config07$theory_n)` sample sizes,
approximately log-spaced from $50$ through $10^7$, including the simulation
sizes. Equation (8) makes this calculation linear in the largest sample size.

The implementation of (3) was checked against a quadratic-form matrix at
$n=2,3,7,19$ for every weight, including the mean against the matrix
trace, the variance against twice the squared Frobenius norm, the score
against the existing increment function, and (8) against (3).
The largest absolute discrepancy is
`r format(max(results07$validation$max_absolute_error), scientific = TRUE, digits = 2)`.
The exact unweighted variance and seeded reproducibility were also checked.

## Do the empirical variances agree with the exact ratios?

Blue curves are exact ratios; orange points use simulated variances divided
by the exact squared mean. The horizontal axis is logarithmic and panels
have separate vertical scales.

```{r ratio-plot, fig.height=6.5, fig.cap="Exact and empirical Chebyshev ratios for the seven main weights."}
ratio_plot07
```

## Do wrong-selection frequencies decrease with sample size?

Points are simulated wrong-selection frequencies and bars are pointwise
95% Wilson intervals. The Monte Carlo standard error is
$\{\widehat q(1-\widehat q)/5000\}^{1/2}$.
A zero count still has a positive Wilson upper endpoint.

```{r probability-plot, fig.height=6.5, fig.cap="Incorrect-selection frequencies and Monte Carlo uncertainty."}
probability_plot07
```

```{r endpoint-table}
kable(endpoint_table07, format = "latex", booktabs = TRUE, digits = 4, label = NA,
      caption = "Exact moments and ratios, with error estimates, at n = 100,000. Lower and Upper are Wilson endpoints.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## Why is the ratio shrinking?

Figure 3 separates mean and variance. For positive log powers both grow,
while for negative log powers the variance has a finite limit.

```{r moment-plot, fig.height=4.5, fig.cap="Theoretical and empirical means and variances."}
moment_plot07
```

## Where is the boundary in the log-power family?

Figure 4 shows the $p=-1$ boundary. Its ratio decreases only as
$(\log\log n)^{-2}$, explaining the slow finite-sample change.

```{r family-plot, fig.height=4.1, fig.cap="Exact ratio trajectories within the log-power family."}
family_plot07
```

For $p<-1$, stopping the exact calculation at a finite $N$ leaves out
a mean tail that shrinks slowly. The omitted part is asymptotic to
$(\log N)^{p+1}/(-p-1)$, even for very large $N$.
Table 3 adds approximate tail corrections and shows how the estimates
change as $N$ grows. It uses

$$
\widehat M_\infty=M_N+\frac{(\log N)^{p+1}}{-p-1},
\qquad
\widehat V_\infty=V_N-2(\log N)^{2p}.
$$

The mean correction has error $O((\log N)^p/N)$ by integral comparison.
For the variance, integration by parts gives
$W_{n-1}=n(\log n)^p\{1-p/\log n+O((\log n)^{-2})\}$.
Substitution into (8) and summation of the tail give
$V_\infty-V_N=-2(\log N)^{2p}
+O((\log N)^{2p-1})$.
The comparisons show numerical stability across these truncations;
the approximations are not certified error bounds.

```{r limit-table}
kable(limit_table07, format = "latex", booktabs = TRUE, digits = 6, label = NA,
      col.names = c("$p$", "$N$", "$\\widehat M_\\infty$",
                    "$\\widehat V_\\infty$", "$\\widehat R_\\infty$"),
      escape = FALSE,
      caption = "Truncation stability of tail-corrected positive-ratio limits.") |>
  kable_styling(latex_options = "hold_position", font_size = 9)
```

\FloatBarrier

## How do these weights compare with a small positive power?

Figure 5 compares three slowly growing weights with $i^{0.01}$ from
Analysis 06.
Its ratio has the positive limit $2(0.01)^2/1.01\approx0.000198$.
The dashed limiting value lies far below the power-weight curve at the
sample sizes shown. The logarithmic ratios eventually fall below that
positive limit because they go to zero, but the figure does not show
where the curves cross. A positive ratio limit for the power weight
still says nothing decisive about its limiting wrong-selection probability.

```{r reference-plot, fig.height=4.1, fig.cap="Slowly varying weights compared with the small positive power from Analysis 06."}
reference_plot07
```

\newpage

# Conclusions by weight family

## Baseline: $w_i=1$

For the unweighted score,

$$
M_n\sim\log n,
\qquad
V_n\to\pi^2,
\qquad
R_n\sim\frac{\pi^2}{(\log n)^2}\to0.
$$

The unweighted baseline is consistent. The mean signal continues to grow
while the variance stays bounded. In the log-power family, $p=0$ is a
special variance regime.

## Increasing log-power weights: $p>0$

For $w_i=[\log(i+1)]^p$ with $p>0$,

$$
M_n\sim\frac{(\log n)^{p+1}}{p+1},
\qquad
V_n\sim2(\log n)^{2p},
\qquad
R_n\sim\frac{2(p+1)^2}{(\log n)^2}\to0.
$$

The tested cases are

$$
p=1:\quad w_i=\log(i+1),
\qquad R_n\sim\frac8{(\log n)^2},
$$

and

$$
p=2:\quad w_i=[\log(i+1)]^2,
\qquad R_n\sim\frac{18}{(\log n)^2}.
$$

Every fixed positive log power tested here is consistent. On the increasing
side of this family, no Chebyshev-ratio consistency boundary appears.

**Rule:** Both mean and variance grow, but the squared mean grows two powers
of $\log n$ faster than the variance.

## Decreasing log-power weights: $p<0$

### Slowly decreasing weights: $-1<p<0$

Here

$$
M_n\sim\frac{(\log n)^{p+1}}{p+1}\to\infty,
\qquad
V_n\to V_\infty(p)<\infty,
\qquad
R_n\to0.
$$

Slowly decreasing weights are still consistent, so increasing weights are
not necessary. One example is

$$
p=-\frac12:\qquad w_i=\frac1{\sqrt{\log(i+1)}}.
$$

### Boundary weight: $p=-1$

For the tested weight $w_i=1/\log(i+1)$,

$$
M_n\sim\log\log n\to\infty,
\qquad
V_n\to V_\infty(-1),
\qquad
R_n\asymp\frac1{(\log\log n)^2}\to0.
$$

Reciprocal log is consistent, although its ratio goes to zero very slowly.

### Faster decreasing weights: $p<-1$

In this range,

$$
M_n\to M_\infty(p)>0,
\qquad
V_n\to V_\infty(p)>0,
\qquad
R_n\to\frac{V_\infty(p)}{M_\infty(p)^2}>0.
$$

The tested example is

$$
p=-2:\qquad w_i=\frac1{[\log(i+1)]^2}.
$$

The total mean signal stops growing, so the Chebyshev ratio no longer
vanishes. This does not prove inconsistency; it only means that the
Chebyshev bound no longer proves consistency.

\pagebreak

Therefore,

$$
\boxed{p=-1\text{ is the Chebyshev-ratio boundary for }[\log(i+1)]^p.}
$$

**Rule:** Decreasing weights can still work if they decrease slowly enough.
The boundary occurs when the mean changes from diverging to converging.

## Polynomial of log

The tested polynomial is

$$
w_i=1+\log(i+1)+[\log(i+1)]^2.
$$

More generally, if

$$
P(x)=cx^k+\text{lower-order terms},
\qquad c>0,\quad k\geq1,
$$

then $P(\log(i+1))\sim c[\log(i+1)]^k$. Consequently,

$$
M_n\sim\frac{c}{k+1}(\log n)^{k+1},
\qquad
V_n\sim2c^2(\log n)^{2k},
\qquad
R_n\sim\frac{2(k+1)^2}{(\log n)^2}\to0.
$$

For the tested degree-two polynomial, $k=2$ and $c=1$, so
$R_n\sim18/(\log n)^2$. The tested polynomial is consistent.

**Rule:** A positive polynomial of $\log i$ is controlled asymptotically by
its highest-degree term. A degree-$k$ polynomial behaves like the log-power
case $p=k$.

## Log-log weight

For $w_i=\log\log(i+e)$,

$$
M_n\sim(\log n)\log\log n,
\qquad
V_n\sim2(\log\log n)^2,
\qquad
R_n\sim\frac2{(\log n)^2}\to0.
$$

The log-log weight is consistent even though it grows more slowly than
$\log i$.

**Rule:** Very slow increasing growth can still be enough; strongly
increasing weights are not necessary.

## Overall rules from Analysis 07

1. For $w_i=[\log(i+1)]^p$, the Chebyshev-ratio boundary is $p=-1$.
2. Every fixed $p>0$ considered here gives consistency.
3. The range $-1\leq p<0$ is also consistent, so weights need not increase.
4. For $p<-1$, both mean and variance converge to positive constants and the
   ratio has a positive limit. The Chebyshev argument is then inconclusive.
5. A polynomial $P(\log(i+1))$ is controlled by its highest-degree term:
   $P(\log(i+1))\sim c[\log(i+1)]^k$. Thus degree $k$ behaves like $p=k$.
6. The log-log weight is consistent, with
   $R_n\sim2/(\log n)^2\to0$.

The main lesson is that consistency is not determined simply by whether the
weights increase or decrease. In this Chebyshev argument, what matters is
whether the accumulated mean signal becomes large enough relative to the
score's standard deviation.
