---
title: "Increasing Logarithmic Weights for Rolling Validation: Simulation Study"
author: "Aileen Tao"
execute:
  dir: project
format:
  pdf:
    number-sections: true
    fig-width: 7.2
    fig-height: 4.6
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{placeins}
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
               fig.pos = "H", out.width = "100%")
here::i_am("report/08B_increasing_logarithmic_weights_simulation.qmd")

results08 <- readRDS(here::here("data", "final", "07_slowly_varying_rv_results.rds"))
weight_ids08 <- c("logp_0", "logp_1", "logp_2", "poly2", "loglog")
weight_labels08 <- c(
  "Unweighted: 1",
  "Log: log(i+1)",
  "Squared log: [log(i+1)]^2",
  "Polynomial log",
  "Log-log: log log(i+e)"
)
weight_key08 <- data.frame(weight_id = weight_ids08, label08 = weight_labels08)

exact08 <- results08$exact |>
  filter(weight_id %in% weight_ids08) |>
  left_join(weight_key08, by = "weight_id") |>
  mutate(label08 = factor(label08, levels = weight_labels08))

mc08 <- results08$mc |>
  filter(weight_id %in% weight_ids08) |>
  left_join(results08$exact, by = c("weight_id", "n")) |>
  left_join(weight_key08, by = "weight_id") |>
  mutate(label08 = factor(label08, levels = weight_labels08),
         empirical_ratio = variance_mc / mean^2)

style08 <- theme_bw(base_size = 10) +
  theme(legend.position = "bottom", strip.text = element_text(size = 8.5),
        panel.grid.minor = element_blank(), plot.caption = element_text(hjust = 0))

mean_plot08 <- ggplot(exact08, aes(n, mean)) +
  geom_line(colour = "#2166AC", linewidth = 0.55) +
  geom_point(data = mc08, aes(y = mean_mc), colour = "#B35806", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sample size n (log scale)", y = "Mean (log scale)",
       caption = "Blue lines: exact expectations. Orange points: Monte Carlo means.") + style08

variance_plot08 <- ggplot(exact08, aes(n, variance)) +
  geom_line(colour = "#2166AC", linewidth = 0.55) +
  geom_point(data = mc08, aes(y = variance_mc), colour = "#B35806", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sample size n (log scale)", y = "Variance (log scale)",
       caption = "Blue lines: exact variances. Orange points: Monte Carlo variances.") + style08

ratio_plot08 <- ggplot(exact08, aes(n, ratio)) +
  geom_line(colour = "#2166AC", linewidth = 0.55) +
  geom_point(data = mc08, aes(y = empirical_ratio), colour = "#B35806", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sample size n (log scale)", y = "Chebyshev ratio (log scale)",
       caption = paste0("Blue lines: exact variance / exact mean squared. ",
                        "Orange points: empirical variance / exact mean squared.")) + style08

probability_plot08 <- ggplot(mc08, aes(n, p_wrong)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.07, colour = "#636363") +
  geom_line(colour = "#2166AC", linewidth = 0.45) +
  geom_point(colour = "#2166AC", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Sample size n (log scale)", y = "Estimated P(Xi[n] < 0)",
       caption = paste0("Pointwise 95% Wilson intervals from 5,000 replications. ",
                        "Shared paths make estimates across weights and n dependent.")) + style08

endpoint08 <- mc08 |>
  filter(n == max(results08$config$mc_n)) |>
  transmute(Weight = as.character(label08),
            `Exact mean` = mean, `MC mean` = mean_mc,
            `Exact variance` = variance, `MC variance` = variance_mc,
            `Exact ratio` = ratio, `Wrong-selection frequency` = p_wrong,
            `Wilson lower` = lower, `Wilson upper` = upper)
```

# Purpose

The companion theory report, *Increasing Logarithmic Weights for Rolling
Validation: Theory* (08A), derives asymptotic expressions for the expectation,
variance, and Chebyshev ratio of several increasing logarithmic weighting
schemes. This report examines whether those patterns are visible at finite
sample sizes and whether the empirical wrong-selection probability decreases
accordingly.

For $Y_i=\theta_\star+\xi_i$ with independent standard Gaussian
$\xi_i$, 08A compares the pre-validation sample mean with the oracle
$\theta_\star$. Write $\Xi_n$ for the weighted difference between their
squared validation losses. A positive value favors the oracle, whereas
$\Xi_n<0$ is an incorrect selection. The weights appear throughout in the
same order:

1. $1$;
2. $\log(i+1)$;
3. $[\log(i+1)]^2$;
4. $1+\log(i+1)+[\log(i+1)]^2$;
5. $\log\log(i+e)$.

The unweighted choice is a reference case. The remaining four weights
increase, at different rates.

# Simulation design

The validated cached experiment contains 5,000 independent replications,
seed 2026, processed in batches of 250. The
sample sizes are
$100,200,500,1000,2000,5000,10000,20000,50000,100000$.
Each Gaussian path is reused across weights and sample sizes, so comparisons
are dependent. Exact moment trajectories extend from $n=50$ through
$10^7$ on an approximately log-spaced grid that includes the simulation
sizes.

The original implementation checks the quadratic-form representation, exact
moments, direct score calculations, and the variance recurrence. The largest
absolute discrepancy across these checks is
`r format(max(results08$validation$max_absolute_error), scientific = TRUE, digits = 2)`.
This report reads those cached results for the five weights above. The
design fixes the comparison before examining the moment, ratio, and selection
plots.

```{r design-table}
design08 <- data.frame(
  Item = c("Replications", "Seed", "Batch size", "Simulation sample sizes",
           "Exact-trajectory range", "Path reuse"),
  Value = c("5,000", "2026", "250",
            "100 to 100,000 (10 specified sizes)",
            "50 to 10,000,000", "Across weights and sample sizes")
)
kable(design08, format = "latex", booktabs = TRUE, label = NA,
      caption = "Simulation design for the five weights.") |>
  kable_styling(latex_options = "hold_position")
```

# Validation of theoretical moments

Before interpreting selection errors, we check whether simulated moments
match the exact moments used in 08A. Separate vertical scales accommodate
the different magnitudes of the five weights.

## Expectation

08A predicts mean growth of order $\log n$ for the baseline,
$(\log n)^2$ and $(\log n)^3$ for the two log powers,
$(\log n)^3$ for the polynomial, and
$(\log n)\log\log n$ for the log-log weight. Figure 1 compares exact
expectations with Monte Carlo means.

```{r mean-plot, fig.height=4.8, fig.cap="Exact and empirical expectations for the five weights."}
mean_plot08
```

The empirical points follow the exact curves across the simulated sizes.
In particular, the polynomial and squared-log means approach the same
leading scale, as the highest-degree-term result predicts.

## Variance

08A predicts variance tending to $\pi^2$ for the baseline, growing as
$2(\log n)^2$ and $2(\log n)^4$ for the log powers and polynomial, and
as $2(\log\log n)^2$ for the log-log weight. Figure 2 compares those exact
variance trajectories with empirical variances.

```{r variance-plot, fig.height=4.8, fig.cap="Exact and empirical variances for the five weights."}
variance_plot08
```

The empirical variances fluctuate more than the means but track the exact
curves. The squared-log and polynomial cases show much larger variability
over this range, even though their leading asymptotic forms agree. These
moment checks support the calculations used in 08A and lead to the ratio
comparison below.

# Chebyshev ratio

08A predicts a ratio asymptotic to a weight-specific constant divided by
$(\log n)^2$ for each choice. Figure 3 examines whether that downward
pattern is visible at finite $n$. The blue curves use exact moments; the
orange points use the Monte Carlo variance and exact mean.

```{r ratio-plot, fig.height=4.8, fig.cap="Exact and empirical Chebyshev ratios."}
ratio_plot08
```

Every exact ratio decreases over the displayed range, and the empirical
points follow the same broad pattern. This is consistent with the predicted
$(\log n)^{-2}$ asymptotic rate, though these finite curves alone do not
establish that rate. The squared-log and polynomial-log curves remain
different at finite $n$ because the polynomial contains lower-degree terms.
The log-log ratio also decreases despite its very slow weight growth.

The ratio is a sufficient bound for consistency; the next section examines
the actual selection errors separately.

# Wrong-selection probability

08A shows that the Chebyshev ratio tends to zero, which is sufficient for
$\mathbb P(\Xi_n<0)\to0$. Figure 4 instead estimates that probability
directly from the simulated fraction of incorrect selections. The ratio is
a theoretical upper-bound quantity; the frequency measures actual
finite-sample errors. Bars are pointwise 95% Wilson intervals.

```{r probability-plot, fig.height=4.8, fig.cap="Wrong-selection frequencies and pointwise Monte Carlo uncertainty."}
probability_plot08
```

Wrong-selection frequencies decline for every weight, as the consistency
results suggest. The rate of finite-sample improvement differs: the decline
is slowest for the squared-log and polynomial-log choices over this range,
consistent with their larger variability. The log and log-log frequencies
also fall, while the unweighted case is a low-error reference. Thus faster
weight growth does not automatically improve finite-sample selection.

```{r endpoint-table}
kable(endpoint08, format = "latex", booktabs = TRUE, digits = 4, label = NA,
      caption = "Exact moments, Monte Carlo moments, ratios, and wrong-selection frequencies at n = 100,000.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 7.5)
```

\FloatBarrier

# Weight-specific interpretation

The moment, ratio, and error plots can now be read together for each
increasing weight.

**Unweighted baseline.** The variance approaches a finite limit, and the
wrong-selection frequency declines across the simulated sizes.

**Log weight.** Its exact mean and variance grow on the scales predicted in
08A, while both the Chebyshev ratio and wrong-selection frequency decline.

**Squared-log weight.** The empirical moments follow the exact trajectories.
The ratio and error frequency decrease more slowly at the available sample
sizes than for the log weight, but the direction agrees with the theory.

**Polynomial-log weight.** Its trajectories resemble those of squared log,
while retaining visible finite-sample differences from the lower-degree
terms. This is the expected approach to highest-degree-term dominance.

**Log-log weight.** Both the ratio and wrong-selection frequency decline even
though the weight grows extremely slowly, matching the qualitative conclusion
of 08A.

# Summary

Across the five weights, the empirical moments track the exact calculations,
the ratios decline in the direction predicted by 08A, and wrong-selection
frequencies decrease over the simulated sizes. The squared-log and
polynomial-log cases improve more slowly at finite $n$, while the log-log
case also shows the predicted downward behavior.

These numerical results do not replace the asymptotic proofs in 08A. They
show how the theoretical mechanisms appear at finite sample sizes and preserve
the distinction between a sufficient Chebyshev argument and the directly
observed wrong-selection frequency.
