---
title: "Increasing Logarithmic Weights for Rolling Validation: Simulation Study"
author: "Aileen Tao"
execute:
  dir: project
format:
  pdf:
    fig-width: 7.2
    fig-height: 4.6
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{placeins}
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
               fig.pos = "H", out.width = "100%")
here::i_am("report/08B_increasing_logarithmic_weights_simulation.qmd")

results08 <- readRDS(here::here("data", "final", "07_slowly_varying_rv_results.rds"))
weight_ids08 <- c("logp_0", "logp_1", "logp_2", "poly2", "loglog")
weight_labels08 <- c(
  "Unweighted: 1",
  "Log: log(i+1)",
  "Squared log: [log(i+1)]^2",
  "Polynomial log",
  "Log-log: log log(i+e)"
)
weight_key08 <- data.frame(weight_id = weight_ids08, label08 = weight_labels08)

exact08 <- results08$exact |>
  filter(weight_id %in% weight_ids08) |>
  left_join(weight_key08, by = "weight_id") |>
  mutate(label08 = factor(label08, levels = weight_labels08))

mc08 <- results08$mc |>
  filter(weight_id %in% weight_ids08) |>
  left_join(results08$exact, by = c("weight_id", "n")) |>
  left_join(weight_key08, by = "weight_id") |>
  mutate(label08 = factor(label08, levels = weight_labels08),
         empirical_ratio = variance_mc / mean^2)

style08 <- theme_bw(base_size = 10) +
  theme(legend.position = "bottom", strip.text = element_text(size = 8.5),
        panel.grid.minor = element_blank(), plot.caption = element_text(hjust = 0))

mean_plot08 <- ggplot(exact08, aes(n, mean)) +
  geom_line(colour = "#2166AC", linewidth = 0.55) +
  geom_point(data = mc08, aes(y = mean_mc), colour = "#B35806", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sample size n (log scale)", y = "Mean (log scale)",
       caption = "Blue lines: exact expectations. Orange points: Monte Carlo means.") + style08

variance_plot08 <- ggplot(exact08, aes(n, variance)) +
  geom_line(colour = "#2166AC", linewidth = 0.55) +
  geom_point(data = mc08, aes(y = variance_mc), colour = "#B35806", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sample size n (log scale)", y = "Variance (log scale)",
       caption = "Blue lines: exact variances. Orange points: Monte Carlo variances.") + style08

ratio_plot08 <- ggplot(exact08, aes(n, ratio)) +
  geom_line(colour = "#2166AC", linewidth = 0.55) +
  geom_point(data = mc08, aes(y = empirical_ratio), colour = "#B35806", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sample size n (log scale)", y = "Chebyshev ratio (log scale)",
       caption = paste0("Blue lines: exact variance / exact mean squared. ",
                        "Orange points: empirical variance / exact mean squared.")) + style08

probability_plot08 <- ggplot(mc08, aes(n, p_wrong)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.07, colour = "#636363") +
  geom_line(colour = "#2166AC", linewidth = 0.45) +
  geom_point(colour = "#2166AC", size = 1.15) +
  facet_wrap(~label08, ncol = 3, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Sample size n (log scale)", y = "Estimated P(Xi[n] < 0)",
       caption = paste0("Pointwise 95% Wilson intervals from 5,000 replications. ",
                        "Shared paths make estimates across weights and n dependent.")) + style08

endpoint08 <- mc08 |>
  filter(n == max(results08$config$mc_n)) |>
  transmute(Weight = as.character(label08),
            `Exact mean` = mean, `MC mean` = mean_mc,
            `Exact variance` = variance, `MC variance` = variance_mc,
            `Exact ratio` = ratio, `Wrong-selection frequency` = p_wrong,
            `Wilson lower` = lower, `Wilson upper` = upper)
```

# Purpose

The companion theory report, *Increasing Logarithmic Weights for Rolling
Validation: Theory* (08A), derives asymptotic expressions for the expectation,
variance, and Chebyshev ratio of several increasing logarithmic weighting
schemes. This report examines whether those patterns are visible at finite
sample sizes and whether the empirical wrong-selection probability decreases
accordingly.

The notation and sign convention match 08A. The statistic is denoted by
$\Xi_n$; a positive value favors the oracle, and $\Xi_n<0$ is incorrect
selection. The weights appear throughout in the same order:

1. $1$;
2. $\log(i+1)$;
3. $[\log(i+1)]^2$;
4. $1+\log(i+1)+[\log(i+1)]^2$;
5. $\log\log(i+e)$.

# Simulation design

The simulation design and cached results are inherited from the authoritative
07 analysis, with the decreasing weights removed from this report. There are
5,000 independent replications, seed 2026, processed in batches of 250. The
sample sizes are
$100,200,500,1000,2000,5000,10000,20000,50000,100000$.
Each Gaussian path is reused across weights and sample sizes, so comparisons
are dependent. Exact moment trajectories extend from $n=50$ through
$10^7$ on an approximately log-spaced grid that includes the simulation
sizes.

The original implementation checks the quadratic-form representation, exact
moments, direct score calculations, and the variance recurrence. The largest
absolute discrepancy across these checks is
`r format(max(results08$validation$max_absolute_error), scientific = TRUE, digits = 2)`.
No Monte Carlo experiment is rerun here; 08B reads the validated 07 cache and
filters it to the five weights above.

```{r design-table}
design08 <- data.frame(
  Item = c("Replications", "Seed", "Batch size", "Simulation sample sizes",
           "Exact-trajectory range", "Path reuse"),
  Value = c("5,000", "2026", "250",
            "100 to 100,000 (10 specified sizes)",
            "50 to 10,000,000", "Across weights and sample sizes")
)
kable(design08, format = "latex", booktabs = TRUE,
      caption = "Simulation design inherited from the authoritative 07 analysis.") |>
  kable_styling(latex_options = "hold_position")
```

# Do empirical moments agree with theory?

The first check compares exact moments with Monte Carlo moments. Agreement
supports both the theoretical formulas used in 08A and their numerical
implementation. Separate vertical scales are used because the five weights
produce moments of very different magnitudes.

```{r mean-plot, fig.height=4.8, fig.cap="Exact and empirical expectations for the five weights."}
mean_plot08
```

```{r variance-plot, fig.height=4.8, fig.cap="Exact and empirical variances for the five weights."}
variance_plot08
```

Across the displayed sample sizes, the empirical means closely follow the
exact expectations. Empirical variances fluctuate more because they estimate
a second moment, but they track the exact variance trajectories. The
polynomial-log moments grow on the same leading scale as the squared-log
moments, as predicted by the highest-degree-term result in 08A.

# Does the Chebyshev ratio follow the prediction?

The theory in 08A predicts a vanishing Chebyshev ratio for all five weights.
The blue curves below are the exact ratios. The orange points replace the
exact variance by the Monte Carlo variance while retaining the exact mean in
the denominator.

```{r ratio-plot, fig.height=4.8, fig.cap="Exact and empirical Chebyshev ratios."}
ratio_plot08
```

Every exact ratio decreases over the displayed range. The empirical points
follow the same broad pattern. The squared-log and polynomial-log curves are
similar but not identical at finite $n$; their agreement is asymptotic because
the polynomial contains lower-degree terms. The log-log ratio also decreases,
despite the exceptionally slow growth of its weights.

# Does the wrong-selection frequency decrease?

The wrong-selection frequency estimates the actual finite-sample probability
$\mathbb P(\Xi_n<0)$. It is not the Chebyshev ratio. The ratio supplies a
sufficient upper-bound argument for asymptotic consistency, whereas the
frequency below directly records observed selection mistakes. Bars are
pointwise 95% Wilson intervals.

```{r probability-plot, fig.height=4.8, fig.cap="Wrong-selection frequencies and pointwise Monte Carlo uncertainty."}
probability_plot08
```

Wrong-selection frequencies decline for every weight. The decline is slowest
for the squared-log and polynomial-log choices over the simulated range,
which is compatible with their larger finite-sample variability. The log and
log-log weights also show decreasing error frequencies, while the unweighted
case serves as a low-error reference.

```{r endpoint-table}
kable(endpoint08, format = "latex", booktabs = TRUE, digits = 4,
      caption = "Exact moments, Monte Carlo moments, ratios, and wrong-selection frequencies at n = 100,000.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 7.5)
```

\FloatBarrier

# Weight-specific interpretation

**Log weight.** The exact mean and variance grow on the scales predicted in
08A, and both the Chebyshev ratio and wrong-selection frequency decline.

**Squared-log weight.** The empirical moments follow the exact trajectories.
The ratio and error frequency decrease more slowly at the available sample
sizes than for the log weight, but the direction agrees with the theory.

**Polynomial-log weight.** Its trajectories resemble those of squared log,
while retaining visible finite-sample differences from the lower-degree
terms. This is the expected approach to highest-degree-term dominance.

**Log-log weight.** Both the ratio and wrong-selection frequency decline even
though the weight grows extremely slowly, matching the qualitative conclusion
of 08A.

# Summary

The simulations support the theoretical moment calculations for all five
weights. Exact and empirical Chebyshev ratios decline, and the observed
wrong-selection frequencies decrease across the simulated sample sizes. The
squared-log and polynomial-log cases show slower finite-sample improvement,
while the log-log case already exhibits the predicted downward behavior.

These numerical results do not replace the asymptotic proofs in 08A. They
show how the theoretical mechanisms appear at finite sample sizes and preserve
the distinction between a sufficient Chebyshev argument and the directly
observed wrong-selection frequency.
