---
title: "Crossing-Point Analysis for Weighted Rolling Validation: Simulation Study"
author: "Aileen Tao"
execute:
  dir: project
format:
  pdf:
    number-sections: true
    fig-width: 7.2
    fig-height: 4.6
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{placeins}
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.pos = "H", out.width = "100%")
here::i_am("report/09_crossing_point_simulation.qmd")
source(here::here("code", "09_crossing_point_simulation.R"),
       local = knitr::knit_global())
```

# Purpose and new model

This report checks the new Analysis 09 crossing formulas at finite training
sizes. The theory report supplies the proofs. The model is
$Y_i\stackrel{\mathrm{iid}}{\sim}N(\theta,1)$, with
$0<|\theta|<1$: Model 0 uses the sample mean of the available training
data, while Model 1 always predicts zero. Model 1 is not an oracle here.
Negative loss differences favor Model 0.

The population comparison after $n$ training observations is
$\Delta_R(n)=1/n-\theta^2$. Expected weighted rolling validation (RV)
has the same sign as
$$
\frac{\Delta_w(n)}{W_n}
=\frac{\sum_{i=2}^{n+1}w_i/(i-1)}
       {\sum_{i=2}^{n+1}w_i}-\theta^2.
$$
The division by $W_n>0$ puts the curves on a comparable scale without
changing a crossing. Ordinary RV is the unweighted reference; log,
squared-log, and log-log weights come from 08; geometric weighting is a new
09 diagnostic for crossing alignment. The order is fixed throughout.

# Design and checks

Exact formulas are evaluated for
$|\theta|\in\{0.2,0.1,0.05,0.025\}$, with training sizes through 40,000.
The geometric base is 1.05. At each population crossing, we examine
$n/n_{\mathrm{pop}}$ near $0.75,1,1.25,2$. The source checks the direct
squared-loss score against a Gaussian quadratic form, and checks its exact
mean and variance; the largest discrepancy is
`r format(results09$validation_error, scientific = TRUE, digits = 2)`.

A separate Monte Carlo check uses 2,000 Gaussian paths at $\theta=0.1$,
seed 2026, in batches of 200. Each path is reused across weights and
training sizes. The code caches its output, so re-rendering does not repeat
the experiment. Exact curves and Monte Carlo points are distinguished in
every figure.

```{r design-table}
design09 <- data.frame(
  Item = c("Gaussian mean magnitudes", "Exact maximum training size",
           "Geometric base", "Monte Carlo theta", "Monte Carlo replications",
           "Seed", "Batch size"),
  Value = c("0.2, 0.1, 0.05, 0.025", "40,000", "1.05", "0.1",
            "2,000", "2026", "200")
)
kable(design09, format = "latex", booktabs = TRUE, label = NA,
      caption = "Analysis 09 calculation and simulation design.") |>
  kable_styling(latex_options = "hold_position")
```

# Do expected RV and population risk cross together?

The theory predicts that ordinary RV crosses late. Logarithmic weighting
should shorten this lag without eliminating it asymptotically; geometric
weighting should move the expected RV crossing close to the population
crossing. Figure 1 compares the exact population and normalized expected RV
differences at $\theta=0.1$. The grey dashed line marks the population
crossing; the orange dotted line marks the expected RV crossing.

```{r comparison-plot, fig.height=4.8, fig.cap="Exact population and expected weighted-RV loss differences at theta = 0.1. Vertical lines mark their crossings."}
comparison09
```

Every weighted crossing is later than the population crossing. The log and
squared-log weights cross earlier than ordinary RV, while geometric
weighting is nearest at this parameter value. This is a crossing comparison,
not a claim about random selection accuracy.

Figure 2 tests the asymptotic direction by shrinking $|\theta|$. The
crossing ratio is one for exact alignment; a ratio above one measures lag.

```{r crossings-plot, fig.height=4.5, fig.cap="Exact expected-RV crossing ratios as the population crossing moves outward."}
crossings_plot09
```

The geometric ratio approaches one over this grid. The logarithmic ratios
rise, consistent with their theoretical logarithmic divergence. Faster
growth within the two log-power examples reduces the lag's constant, not
its asymptotic order.

```{r crossings-table}
crossing_display09 <- crossing09 |>
  mutate(Weight = as.character(label),
         Improvement = ifelse(weight == "unweighted", "Reference",
                              ifelse(improves_over_unweighted, "Yes", "No"))) |>
  select(`|theta|` = theta, Weight, `Population crossing` = n_pop,
         `Expected RV crossing` = n_rv,
         `Absolute difference` = difference,
         `Crossing ratio` = crossing_ratio, Improvement)
kable(crossing_display09, format = "latex", booktabs = TRUE,
      label = NA, digits = 3,
      caption = "Exact crossing indices and improvement relative to ordinary RV.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 7.5)
```

\FloatBarrier

# Does alignment retain enough averaging?

Analysis 09 proves that, near the moving crossing, the sufficient
alignment condition $nA_n/W_n\to1$ forces
$n\sum w_i^2/W_n^2\to\infty$. The new linear Gaussian variance term
then prevents a vanishing Chebyshev ratio on either fixed side
$n/n_{\mathrm{pop}}\to c\ne1$. Figure 3 uses the **exact** 09 variance,
including both quadratic and linear parts, at three values of $c$.
The ratio at $c=1$ is omitted from this plot because the leading
population difference is zero there.

```{r tradeoff-plot, fig.height=4.5, fig.cap="Exact Chebyshev ratios on both sides of the moving population crossing."}
tradeoff09
```

The log weights have small ratios over the shrinking-$\theta$ grid, but
for $c>1$ their expected RV comparison still favors the wrong model.
Geometric weighting approaches crossing alignment, but its ratio stays
large and can be irregular at finite sizes near its own crossing.
This figure illustrates the two distinct failures identified by the
theory; it does not establish a limiting selection probability.

The boundary $c=1$ is shown separately in the table below. Its
finite-grid ratios are descriptive: an asymptotic ratio with an exactly
zero mean would be undefined.

```{r local-table}
local_display09 <- local09 |>
  filter(theta == 0.1, c_target %in% c(0.75, 1, 1.25, 2)) |>
  mutate(Weight = as.character(label)) |>
  select(`Target c` = c_target, Weight, n,
         `Population difference` = delta_pop,
         `Expected RV / W` = delta_expected,
         `Quadratic variance / W^2` = variance_quadratic,
         `Linear variance / W^2` = variance_linear)
kable(local_display09, format = "latex", booktabs = TRUE,
      label = NA, digits = 5,
      caption = "Exact near-crossing means and variance components at theta = 0.1.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 7)
```

\FloatBarrier

# Monte Carlo check: expectation versus realized selection

The same theoretical mean can be checked without treating simulated
selection as a proof. Figure 4 compares exact normalized expected RV
curves with Monte Carlo means from the shared paths.

```{r mc-mean-plot, fig.height=4.8, fig.cap="Exact normalized expected RV differences (blue) and Monte Carlo means (orange) at theta = 0.1."}
mc_mean_plot09
```

The empirical means follow the exact curves. Figure 5 asks a different
question: how often does the realized RV score select Model 0?
Pointwise bars use Monte Carlo standard errors; shared paths make
comparisons across weights and sizes dependent.

```{r mc-select-plot, fig.height=4.8, fig.cap="Monte Carlo frequency of selecting Model 0, with pointwise approximate 95% intervals."}
mc_select_plot09
```

For $n/n_{\mathrm{pop}}>1$, population risk favors Model 0.
Expected sign alignment and a realized selection frequency near one are
different requirements. The plotted frequencies describe this finite
experiment; no asymptotic probability claim is inferred from them.

# What these calculations establish

The exact crossing table and curves verify the finite-index formulas and
illustrate the lag that general weighting changes. The exact variance
calculations display the new linear noise term. Monte Carlo means support
the expected RV calculation, while selection frequencies show the
additional randomness faced by a realized criterion. The proofs and
near-crossing impossibility statement are in the separate 09 theory
report.
