---
title: "Cross-Validation: Linear vs Polynomial Regression on Auto Data"
author: Aileen Tao
format:
  pdf:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Plan

In this project, I want to understand how MPG relates to horsepower in the **ISLR Auto dataset**. A simple way to model this relationship is with a straight line, but the relationship might also be curved. If the pattern bends, a linear model may miss some structure in the data, while a polynomial model could capture it better.

To investigate this, I compare two models that predict **mpg using horsepower**:

-   **Linear: mpg \~ horsepower**
-   **Polynomial (degree 3): mpg \~ poly(horsepower, 3)**

I fixed the polynomial degree at 3 ahead of time instead of searching for the “best” degree using the full dataset. This avoids tuning the model in a way that could bias the evaluation.

To compare the models fairly, I use 10-fold cross-validation. The data are split into 10 folds, the model is trained on 9 folds, and tested on the remaining fold. This process is repeated so every observation is used for validation once.

Because the result of a single split can depend on random chance, I repeat the entire 10-fold CV procedure 50 times using different random partitions of the data. For each run I compute the mean validation MSE, and then compare the averages across the 50 runs.

I also compute the training MSE using the full dataset, mainly to illustrate how training error can look overly optimistic compared to cross-validation.



# Methods and replication

All analysis steps are written so the results can be reproduced exactly.

First, I load the required packages and set a seed so the random sampling is consistent.

```{r packages-and-seed}
library(ISLR)
library(tidyverse)

set.seed(2026)
K <- 10L
n_rep <- 50L
poly_degree <- 3L
```

Then I load the Auto dataset from the ISLR package and inspect its structure. The dataset contains 392 cars and 9 variables, including mpg, horsepower, weight, cylinders, and several other attributes.

```{r dataset-info}
data(Auto, package = "ISLR")
glimpse(Auto)
```

Next, I perform a small amount of cleaning. If horsepower appears as a character variable, I convert it to numeric and treat any problematic values as missing. Rows with missing values are then removed.

```{r load-clean}
data(Auto, package = "ISLR")
if (any(sapply(Auto, is.character))) {
  Auto[] <- lapply(Auto, function(v) {
    if (is.character(v)) utils::type.convert(v, as.is = TRUE) else v
  })
}
if ("horsepower" %in% names(Auto) && is.character(Auto$horsepower)) {
  Auto$horsepower <- suppressWarnings(as.numeric(Auto$horsepower))
}
Auto <- stats::na.omit(Auto)
n <- nrow(Auto)
n
```

After cleaning, the dataset contains 392 observations, with:

-   **mpg** as the response variable
-   **horsepower** as the predictor

I use 10-fold cross-validation and repeat the procedure 50 times. In each repetition, the data are randomly divided into 10 folds. For each fold, the models are trained on the other nine folds and evaluated on the remaining fold. I fit both the linear model and the degree-3 polynomial model on the training data, then predict MPG on the validation fold and compute the mean squared error (MSE). The MSEs from the 10 folds are averaged to produce one cross-validation error for that repetition. Repeating the procedure 50 times gives 50 estimates of the mean CV MSE for each model. I then compute the average of those values, their standard deviation, and the training MSE using the full dataset so the models can be compared using both validation and training error.

```{r cv-loops-repeated}
rep_mean_cv_linear <- numeric(n_rep)
rep_mean_cv_poly <- numeric(n_rep)

for (r in seq_len(n_rep)) {
  fold_id <- sample(rep(seq_len(K), length.out = n))
  rep_mean_cv_linear[r] <- kfold_cv_linear(
    Auto,
    mpg ~ horsepower,
    K = K,
    fold_id = fold_id
  )$mean_cv_mse
  rep_mean_cv_poly[r] <- kfold_cv_polynomial(
    Auto,
    response = "mpg",
    predictor = "horsepower",
    degree = poly_degree,
    K = K,
    fold_id = fold_id
  )$mean_cv_mse
}

mean_cv_mse_linear <- mean(rep_mean_cv_linear)
mean_cv_mse_poly <- mean(rep_mean_cv_poly)
sd_rep_linear <- sd(rep_mean_cv_linear)
sd_rep_poly <- sd(rep_mean_cv_poly)
mean_cv_mse_linear
mean_cv_mse_poly
sd_rep_linear
sd_rep_poly
```

```{r training-mse-full}
fit_lin_full <- lm(mpg ~ horsepower, data = Auto)
fit_poly_full <- lm(
  mpg ~ poly(horsepower, degree = poly_degree, raw = TRUE),
  data = Auto
)

train_mse_linear <- mean(residuals(fit_lin_full)^2)
train_mse_poly <- mean(residuals(fit_poly_full)^2)
train_mse_linear
train_mse_poly
```



# Results

```{r results-table, echo=FALSE}
results_rep <- data.frame(
  repetition = seq_len(n_rep),
  mean_CV_MSE_linear = rep_mean_cv_linear,
  mean_CV_MSE_polynomial = rep_mean_cv_poly
)
knitr::kable(
  results_rep,
  digits = 4,
  caption = "Each row is one full 10-fold CV run: mean validation MSE for that run (same folds for linear vs polynomial within the row)."
)

summary_cv <- data.frame(
  model = c("Linear", paste0("Polynomial (degree ", poly_degree, ")")),
  mean_of_repetition_means = c(mean_cv_mse_linear, mean_cv_mse_poly),
  sd_across_repetitions = c(sd_rep_linear, sd_rep_poly),
  training_MSE_full_data = c(train_mse_linear, train_mse_poly)
)
knitr::kable(
  summary_cv,
  digits = 4,
  caption = "Across 50 repeated 10-fold CVs: average of the 50 run-level mean CV MSEs, SD across those 50 values, and training MSE on full data."
)
```


```{r selection, echo=FALSE}
selected_label <- if (mean_cv_mse_linear <= mean_cv_mse_poly) {
  "linear: mpg ~ horsepower"
} else {
  paste0("polynomial (degree ", poly_degree, "): mpg ~ poly(horsepower, ", poly_degree, ")")
}
```

Averaging across the **50** repeated 10-fold CV runs, the mean CV MSE is **24.2607** for the linear model and **19.3514** for the polynomial model. Since the polynomial model has the smaller validation error, I would choose the **polynomial (degree 3) model: mpg \~ poly(horsepower, 3)**.

```{r optional-figure, fig.cap="Quick visual: I fit both models on all the data just to show the shape of the fits. The actual model choice above comes from cross-validation, not from how nice this plot looks.", fig.width=7, fig.height=4}
x_seq <- seq(min(Auto$horsepower), max(Auto$horsepower), length.out = 200)
newdf <- data.frame(horsepower = x_seq)
pred_line <- predict(fit_lin_full, newdata = newdf)
pred_curve <- predict(fit_poly_full, newdata = newdf)

plot(
  Auto$horsepower,
  Auto$mpg,
  xlab = "Horsepower",
  ylab = "MPG",
  main = "MPG vs horsepower (full-data fits for intuition)"
)
lines(x_seq, pred_line, col = "blue", lwd = 2)
lines(x_seq, pred_curve, col = "red", lwd = 2)
legend(
  "topright",
  legend = c("Linear", paste0("Polynomial deg. ", poly_degree)),
  col = c("blue", "red"),
  lwd = 2
)
```



# Conclusion

In this project, I compared a simple linear model and a cubic polynomial model for predicting MPG using horsepower from the Auto dataset. To make the comparison fair, I used 10-fold cross-validation repeated 50 times, with both models evaluated on the same folds each time.

Looking at the average validation error across the 50 runs, the polynomial model performed noticeably better. The mean CV MSE was about 24.26 for the linear model and 19.35 for the polynomial model, so the polynomial model consistently had lower prediction error. Because cross-validation measures how well a model performs on unseen data, this result suggests that the cubic polynomial captures the relationship between horsepower and MPG more effectively than a straight line.

Training error alone could be misleading here. As expected, the polynomial model also had a lower training MSE, since more flexible models tend to fit the existing data better. That is exactly why cross-validation is important. It helps evaluate how the model is likely to perform on new observations rather than just how well it fits the current dataset.

Overall, based on the repeated cross-validation results, I would choose the degree-3 polynomial model for predicting MPG from horsepower in this dataset. However, it’s important to remember that this analysis only focuses on prediction with one variable. Since the data are observational, this model should not be interpreted as evidence that horsepower directly causes changes in MPG.
